---
title: "Exploring relationships within groups"
author: "Charles Crabtree and Michael J. Nelson"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exploring relationships within groups}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Look within groups

We built `plotrr` to inspect relationships within groups before fitting a model
to pooled data. Exploratory data analysis helps us see the observations and
assumptions we're bringing to an analysis (Tukey 1977). It can also reveal
unusual cases that deserve a closer look (NIST/SEMATECH 2012).

Here we'll use simulated data to show how the plotting functions work. The
patterns come from the simulation; they aren't evidence about a population.

### Relationships within groups

As demonstrated in Crabtree and Nelson (2017), group-level plots can help us
see whether an expected relationship appears in particular cases. We'd use
those patterns to guide further investigation, with the research design still
determining what we can say about causes.

#### `bivarplots` 
`bivarplots()` prints a scatterplot and smooth curve for each group. Check how
many groups you have first with `lengthunique()`; a large dataset can produce
a lot of plots.

```{r, fig.show='hold'}
library(plotrr)
a <- runif(400, min = 0, max = 1)
b <- a + rnorm(400, mean = 0, sd = 1)
c <- rep(c(1:4), times = 100)
data <- data.frame(a, b, c)
bivarplots("a", "b", "c", data)
```

#### `dotplots` 
`dotplots()` prints the distribution of x as stacked dots for each group. Its y argument is retained for compatibility and is unused.
```{r, fig.show='hold'}
library(plotrr)
a <- runif(400, min = 0, max = 1)
b <- a + rnorm(400, mean = 0, sd = 1)
c <- rep(c(1:4), times = 100)
data <- data.frame(a, b, c)
dotplots("a", "b", "c", data, 20)
```

#### `violinplots` 
`violinplots()` shows the distribution of y across values of x within each group.
```{r, fig.show='hold'}
library(plotrr)
a <- runif(400, min = 0, max = 1)
b <- a + rnorm(400, mean = 0, sd = 1)
c <- rep(c(1:4), times = 100)
data <- data.frame(a, b, c)
violinplots("a", "b", "c", data)
```

### Comparing measures

We can use these plots to examine competing measures within groups. Agreement
between measures is one part of assessing construct validity (Trochim & Donnelly, 2008).
A disagreement in a particular case can also suggest a useful question: what
is each measure capturing there?

### Other visualization functions

`histplots()` prints a histogram of x for each group. Set `n` to choose the number of bins; the y argument is unused.

```{r, fig.show='hold'}
library(plotrr)
a <- rnorm(400, mean = 2, sd = 100)
b <- a + rnorm(400, mean = 0, sd = 1)
c <- rep(c(1:4), times = 100)
data <- data.frame(a, b, c)
histplots("a", "b", "c", data, 5)
```

`bivarrugplot()` adds a mark along each axis for every observed value. The rugs help show where observations are concentrated.

```{r, fig.show='hold'}
library(plotrr)
a <- runif(1000, min = 0, max = 1)
b <- a + rnorm(1000, mean = 0, sd =1)
data <- data.frame(a, b)
bivarrugplot("a", "b", data)
```

### Other functions

`lengthunique()` counts distinct non-missing values. `makefacnum()` reads a
factor's labels as numbers; non-numeric labels become missing values with a
warning. `clear()` prints a form-feed character, which clears some console
displays.

#### References
- Crabtree, Charles, and Michael J. Nelson. 2017. "New Evidence for a Positive Relationship Between De Facto Judicial Independence and State Respect for Empowerment Rights." _International Studies Quarterly_.
- NIST/SEMATECH. 2012. _e-Handbook of Statistical Methods_. [https://www.itl.nist.gov/div898/handbook/](https://www.itl.nist.gov/div898/handbook/).
- Trochim, William M. K., and James P. Donnelly. 2008. _Research Methods Knowledge Base_. New York, NY: Cengage Learning.
- Tukey, John W. 1977. _Exploratory Data Analysis_. New York, NY: Pearson.
